Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-03-28 tagless/tagless-final.tex
just some preliminary bib work

We have programmed out interpreters in OCaml/MetaOCaml (cite
www.metaocaml.org) and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.

@@

        Taha et al and Xi et al (both POPL2003) argued that writing an
interpreter of a typed language in a typed language is an interesting
problem. The known solutions include the use of the Universal type, or
GADTs. The former is not quite satisfactory for performance reasons;
and also because of the presence of partial pattern matching in the
interpreter; the inexhaustiveness of the pattern matching should never
be triggered when executing a well-typed term. Alas, this fact cannot
be made obvious to the compiler. That also exlains unoptimality.

@@

Of course there is also the issue how to design a type system to type
terms like |test1| (GADT seem necessary). Also, how to design the
typesystem to keep track of open/closed status of terms, so we can
statically assure that only closed terms are passed to eval and that
the empty list checking in |List.hd| and |List.tl| can be eliminated
as well. The latter is even more interesting problem requiring even
more advanced type systems (cite env classifiers, Nanevski, Rowan
Davies, Pitts).

@@

See the Related Work section (\ref{related}) for a discussion of the other
solutions to this problem.
\oleg{I feel we need to give some details of the other solutions here, 
to give the reader the feeling of
how advanced solutions were proposed and how rarely they are
implemented.}

>>>>

We have programmed out interpreters in OCaml/MetaOCaml (\cite{metaocaml})
and Haskell. The complete code is available 
\footnote{the URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
in the supplemental material to the paper. We will use both
(Meta)OCaml and Haskell for our examples throughout the paper.
Each language has its strength and weaknesses; this dual implementation
forced us to clarify our own ideas.  We will continually leverage the
respective strengths of these languages, as certain of our claims are
patently obvious in one implementation, while requiring more argumentation
in the other.

@@

    At POPL 2003, Taha et al~\cite{TahaPOPL03} and Xi et al~\cite{XiPOPL03}
both
interpreter of a typed language in a typed language is an interesting
problem. The known solutions include the use of the Universal type, or
GADTs. The former is not quite satisfactory for performance reasons;
and also because of the presence of partial pattern matching in the
interpreter; the inexhaustiveness of the pattern matching should never
be triggered when executing a well-typed term. Alas, this fact cannot
be made obvious to the compiler. That also exlains unoptimality.

@@

Of course there is also the issue how to design a type system to type
terms like |test1| (GADT seem necessary). Also, how to design the
typesystem to keep track of open/closed status of terms, so we can
statically assure that only closed terms are passed to eval and that
the empty list checking in |List.hd| and |List.tl| can be eliminated
as well. The latter is even more interesting problem requiring even
more advanced type systems (cite env classifiers, Nanevski, Rowan
Davies, Pitts).
(\cite{WalidPOPLo3,NanevskiICFP02,NanevskiJFP05,DaviesJACM01}
\jacques{I hope that is the right Davies paper}

@@

See the Related Work section (\ref{related}) for a discussion of the other
solutions to this problem.
\oleg{I feel we need to give some details of the other solutions here, 
to give the reader the feeling of
how advanced solutions were proposed and how rarely they are
implemented.}
\jacques{Sure, we can refer to this section from the Related Work section
for details.}
